The Effects of the 2023 Rule Changes in Major League Baseball

### 2023-11-09 manuscript/paper.tex
added a couple paragraphs to data section

Major League Baseball (MLB), often referred to as America's pastime, is a sport
rich in tradition and innovation. Throughout its long history, MLB has enacted
numerous rule changes aimed at improving the game's relevance, competitiveness,
and entertainment value. For instance, in 1973, the designated hitter rule was
implemented for all American League teams, with the hopes of increasing
offensive production as a whole. However, we are not really interested in \emph{why}
the rules are changed, but more the underlying effects of the change(s). This
can be seen in \citet{Cooley} where the effects of the designated hitter rule
on player longevity, pitcher complete games, and number of hit batters are
analyzed. We see many more major and minor MLB rule changes come into action
throughout the course of its history, like the recent \emph{ghost runner} rule
carried out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
automatically starts out on second-base in extra-innings.

@@

These rule changes may seem ridiculous to many baseball \emph{purists} out there,
but here is a quote from Commissioner Manfred that explains the rationale:

@@

Also, we do not need to just look at rule changes specifically, we can explore
a \emph{before and after analysis} of some other change made in sports. \citet{Price}
is a good illustration of this is, as they conducted an analysis of the COVID
\emph{bubble} in the NBA playoffs and how this affected home-court advantage.
Nine different z-tests were conducted to try to see if there was significance in
the difference in free-throw rate, winning percentage, etc. in the 2020 season
compared to previous years. This highly relates to our study, but we will instead
be testing the difference in say stolen bases for example, from the 2023 season
compared to previous years. \citet{Zhang} is another great example as they
focused on the effect of the video assistant referee (VAR) on referee's decisions
in the FIFA Women's World Cup. They used a Mann-Whitney U test to compare differences
in game-related variables such as playing time, penalties, and goals, with and
without VAR.

@@

Use this section to describe the data that helps to answer your research
questions.

>>>>

Major League Baseball (MLB), often referred to as America's pastime, is a sport
rich in tradition and innovation. Throughout its long history, MLB has enacted
numerous rule changes aimed at improving the game's relevance, competitiveness,
and entertainment value. For instance, in 1973, the designated hitter rule was
implemented for all American League teams, with the hopes of increasing
offensive production as a whole. However, we are not really interested in ``why''
the rules are changed, but more the underlying effects of the change(s). This
can be seen in \citet{Cooley} where the effects of the designated hitter rule
on player longevity, pitcher complete games, and number of hit batters are
analyzed. We see many more major and minor MLB rule changes come into action
throughout the course of its history, like the recent ``ghost runner'' rule
carried out in the weird, shortened, COVID-19 season in 2020, where a baserunner 
automatically starts out on second-base in extra-innings.

@@

These rule changes may seem ridiculous to many baseball ``purists'' out there,
but here is a quote from Commissioner Manfred that explains the rationale:

@@

Also, we do not need to just look at rule changes specifically, we can explore
a ``before and after analysis'' of some other change made in sports. \citet{Price}
is a good illustration of this is, as they conducted an analysis of the COVID
``bubble'' in the NBA playoffs and how this affected home-court advantage.
Nine different z-tests were conducted to try to see if there was significance in
the difference in free-throw rate, winning percentage, etc. in the 2020 season
compared to previous years. This highly relates to our study, but we will instead
be testing the difference in say stolen bases for example, from the 2023 season
compared to previous years. \citet{Zhang} is another great example as they
focused on the effect of the video assistant referee (VAR) on referee's decisions
in the FIFA Women's World Cup. They used a Mann-Whitney U test to compare differences
in game-related variables such as playing time, penalties, and goals, with and
without VAR.

@@

The data used for this analysis was collected from \href{https://baseballsavant.mlb.com/}{Baseball Savant},
a comprehensive and interactive platform that contains advanced statistics and
data related to Major League Baseball (MLB). The main variables of interest
are Stolen Bases (SB), Batting average (BA), Strike Outs (SO), and Batting
Average on Balls in Play (BABIP), among some others. These are game-related
baseball statistics that will be analyzed similar to those used in the literature
above, like we saw with Free Throw Rate in the NBA, for example. These variables
were selected based on domain knowledge and intuition about what metrics may be
affected most by the new MLB rule changes. We expect to see Stolen Bases increase
across the league, due to the pitch clock and pitcher disengagement limits. Batting
Average, Batting Average on Balls in Play, and Slugging (SLG) may be affected by
the limit of the defensive shift, as this rule change may increase the likelihood
of a hit. In a similar vein, Strike Outs and Walks (BB) may see an impact as both
pitchers and batters adjust to the pitch clock. Other variables included in the
study:
\begin{itemize}
  \item Ground Into Double Play (GIDP)
  \item Hits on Ground Balls (Hit-GB)
  \item Expected Batting Average (xBA)
\end{itemize}
We expect these variables will be affected by the limit of the defensive shift.
Also, pace of play (game time) data was collected from \href{https://www.baseball-reference.com/}{Baseball Reference}.
Each row of this dataset represents a single season from 2023 dating back to 1998
(26 observations total), containing the Average Game Time (in hours) for that season.
Intuitively, we should expect this variable to have a significant decrease for the
2023 season, due to the introduction of the pitch clock making the games go faster.
Ideally, we would have used datasets containing the game times for \underline{each}
game in both seasons, however this was not publicly available.

The main data used though was collected using the Custom Leaderboard feature
on Baseball Savant, where each variable above was added to the dataset, with each
row representing a single player's overall statistics. The data was then filtered
to only include the 2023 season, and only include players with a minimum 100 Plate Appearances (PA)
that season to limit outliers (100 min. PA is a popular threshold used for 
baseball player analysis). This gives us 461 observations for the 2023 season,
again with each row being a unique batter. The same steps were then applied to
create a comparison dataset now filtering for the 2022 season, which has 469
observations total. This gives us a baseline to compare the variables in the
season where the rule changes were in effect (2023), to the previous season where
they were not (2022).

Different data formats were considered, such as using team-wide statistics
for each season instead of player-wide, where the dataset would be 30 rows (one
for each team). Using a combination of previous seasons data for comparison (not
just 2022) was considered also, but we settled on the data we used as it seemed
the most straightforward for our analysis and limited potential confounding
factors. Data from complete game logs for each game of the 2023 season would have
been an option, however this data is not yet publicly available. More on these topics
are discussed in Section's~\ref{sec:meth} and \ref{sec:disc}.
